## Calculating a 95% confidence interval of a mean: individual data {#sec-cimean-r-ind}

To demonstrate the computation of the 95% confidence interval of a mean we have used data from `mod02_weight_30.rds` which contains the weights of 30 students:

```{r}
students <- readRDS("data/examples/mod02_weight_30.rds")
```

We can examine the data set using the `summary` command:

```{r}
summary(students)
```

The mean and its 95% confidence interval can be obtained many ways in R. We will use the `t.test()` function installed in R to calculate the confidence interval:

```{r}
t.test(students$weight)
```

The output of the `t.test()` function gives us the sample mean (70.0 kg) as well as the 95% confidence interval around the mean: 68.1 to 71.9 kg.

Note: the `descriptives()` function within the `jmv` package also calculates a 95% confidence interval around the mean. **It is recommended not to use this function** as it currently (as of June 2022) uses a *z* value to calculate the confidence interval, rather than a *t* value.


## Calculating a 95% confidence interval of a mean: summarised data {#sec-cimean-r-summ}

For Worked Example 3.2 where we are given the sample mean, sample standard deviation and sample size. R does not have a built-in function to calculate a confidence interval from summarised data, but we can write our own.

**Note: writing your own functions is beyond the scope of this course. You should copy and paste the code provided to do this.**

```{r}
### Copy this section
ci_mean <- function(n, mean, sd, width=0.95, digits=3){
  lcl <- mean - qt(p=(1 - (1-width)/2), df=n-1) * sd/sqrt(n)
  ucl <- mean + qt(p=(1 - (1-width)/2), df=n-1) * sd/sqrt(n)
  
  print(paste0(width*100, "%", " CI: ", format(round(lcl, digits=digits), nsmall = digits),
               " to ", format(round(ucl, digits=digits), nsmall = digits) ))

}
### End of copy

ci_mean(n=242, mean=128.4, sd=19.56, width=0.95)
ci_mean(n=242, mean=128.4, sd=19.56, width=0.99)
```
