# Correlation and simple linear regression

```{r}
library(ggformula)
```

We will demonstrate using R for correlation and simple linear regression using the dataset `Example_8.1.csv`.

```{r}
lung <- read.csv("data/examples/Example_8.1.csv")
```

## Creating a scatter plot

We can use the `plot` function to create a scatter plot to explore the association between height and FVC, assigning meaningful labels with the `xlab` and `ylab` commands:

```{r}
plot(x=lung$Height, y=lung$FVC, 
     xlab="Height (cm)", 
     ylab="Forced vital capacity (L)")
```

To add a fitted line, we can use the `abline()` function which adds a straight line to the plot. The equation of this straight line will be determined from the estimated regression line, which we specify with the `lm()` function, which fits a *linear model*.

The basic syntax of the `lm()` function is: `lm(y ~ x)` where `y` represents the *outcome* variable, and `x` represents the *explanatory* variable. Putting this all together:

```{r}
plot(x=lung$Height, y=lung$FVC,
     xlab="Height (cm)",
     ylab="Forced vital capacity (L)")

abline(lm(lung$FVC ~ lung$Height))
```

Or using the `ggformula` package, we form the basic plot using the following:
```{r}
gf_point(FVC ~ Height, data=lung,
     xlab="Height (cm)", 
     ylab="Forced vital capacity (L)") |>
  gf_theme(theme = theme_minimal())
```

We can add an estimated linear regression line by piping the command `gf_lm()`:
```{r}
gf_point(FVC ~ Height, data=lung,
     xlab="Height (cm)", 
     ylab="Forced vital capacity (L)") |>
  gf_lm() |>
  gf_theme(theme = theme_minimal())
```

### Calculating a correlation coefficient {-}

We can use the `cor.test(x, y)` function to calculate a Pearson's correlation coefficient:

```{r}
cor.test(lung$Height, lung$FVC)
```

## Fitting a simple linear regression model

We can use the `lm` function to fit a simple linear regression model, specifying the model as `y ~ x` where `y` represents the *outcome* variable, and `x` represents the *explanatory* variable. Using `Example_8.1.rds`, we can quantify the relationship between FVC and height:

```{r}
lm(lung$FVC ~ lung$Height)
```

The default output from the `lm` function is rather sparse. We can obtain much more useful information by defining the linear regression model as an object, then using the `summary()` function:

```{r}
model <- lm(lung$FVC ~ lung$Height)
summary(model)
```

Finally, we can obtain 95% confidence intervals for the regression coefficients using the `confint` function:

```{r}
confint(model)
```

## Plotting residuals from a simple linear regression

We can use the `resid` function to obtain the residuals from a saved model. These residuals can then be plotted using a histogram in the usual way:

```{r}
residuals <- resid(model)
hist(residuals)
```

A Normal curve can be overlaid if we plot the residuals using a probability scale.

```{r}
hist(residuals, probability = TRUE,
     ylim = c(0, 1))

curve(dnorm(x, mean=mean(residuals), sd=sd(residuals)), 
      col="darkblue", lwd=2, add=TRUE)
```

Using `ggformula`, we can plot the residuals as a histogram:

```{r}
gf_dhistogram(~ residuals, data=model) |>
  gf_dist("norm", 
          params=list(mean=mean(model$residuals), 
                      sd=sd(model$residuals)))

gf_dhistogram(~ residuals, data=model) |>
  gf_dens()
```
